A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-26 paper.tex
Move more text around and trim down

\section*{Node vs edge annotations}
Another aspect of the classical description of an ARG data structure that
derives from its historical origins as a stochastic
process is how we encode information about genetic
inheritance in the graph. The approach taken by Griffiths to
doing this is concise, elegant: all we need to do is store
the location of the breakpoint at each recombination event.
Storing a breakpoint for each recombinatin event is a very
natural and straightforward thing to do when simulating
a (``Big''; see Appendix XXX) ARG process.
Given the graph topology and these breakpoints, we can then
recover the
local trees at every point in the genome
by tracing rootwards from the leaves and choosing the
appropiate outbound path at recombination nodes.
However, there are several significant disadvantages
to this approach: [TODO be concrete here later
and give pointers to the section where we discuss the
alternative.]

Fortunately, there is a straightforward change to the
classical eARG encoding that simplifies
implementations and greatly increases the flexibility
of both the types and granularity of inheritance information
that we can represent.
The main change that we need to make is to
associate the
the details of inheritance with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules.
This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

\section*{Genome ARGs}
As we have seen in previous sections, the focus
on a limited palette of events makes describing currently-available
datasets and known patterns of genetic inheritance as an eARG
quite cumbersome.
There is no fundamental reason that the patterns of genetic inheritance
that we wish to describe via an ARG must be in terms
of discrete \emph{events}, however. Recent discussions have illustrated the
relationship between ARGs and
the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
and the deep connections between the structures are
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}.
Following \cite{mathieson2020ancestry},
rather defining the inheritance
of genetic material in terms of \emph{events} as we do in an eARG,
we can instead define an ARG data structure in terms of
\emph{genomes} and intervals of genetic inheritance
along the edges (see the previous section).
To distinguish from the classical Griffiths eARG,
we refer to this encoding of an ARG data structure
as a ``Genome ARG'', or gARG.
The distinction between nodes representing events
and genomes may seem pedantic at first, but as we
see in later sections, we gain a great deal of flexibility
and computational efficiency by decoupling our representation
of genetic ancestry from the events that generated it.

@@

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated (see the previous section) with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.

Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity what can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome.
In large deeply sequenced pedigrees such complexities are commonplace.

Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can intepret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes (see section XXX).
Whether one regards nodes as genomes or events is orthogonal,
and something of a philosophical question which makes little
difference from a computational perspective.

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
parent to child (and ultimately from cell to cell) through some
number of generations. The edges are ``annotated'' (see section XXX)
with the coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to define the annotations as
the genetic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.

@@

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlike to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).

@@

\section*{Precision of ancestry information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genetic history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information
will always be available, and in particular whether our
representation of genealogical history should \emph{require}
such precision. Simulations will usually generate complete
information about the history of a sample, of course, but
we may not have sufficient information to infer such detail.

@@

To illustrate this critical point, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
unrelated local trees
% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% for discussion. I'm sure there were other examples of people saying
% that
(as claimed by e.g.~\citep{hejase2020summary}),
and information about recombination and the correlation structure
between trees is not ``all or nothing''.
Both methods capture correlation structure between trees:
\relate\ first infers local tree topologies and then identifies edges
that persist from one tree to the next as a separate step
(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
\texttt{tsinfer} explicitly constructs nodes that are shared across local
trees.

@@

\section*{ARG simplification}\label{ARG_simplification}

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow of ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the ``meaningful'' topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
neither could be directly represented in a Griffiths-like eARG encoding
(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

% Q: would it help to talk about ``tree vertices'' here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g.~node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, \emph{all}
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
It is common for this operation to remove nodes above the top MRCA node
in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
become roots). [And continue]
% Note that although this last stage produces much simpler local trees, by
% removing information about the exact paths taken by lineages through
% the graph, it loses potentially important information about shared edges
% between trees. It is not clear whether (c) or (d) leads to a more compact
% representation of the ARG in terms of number of edges.

>>>>

% \section*{Node vs edge annotations}
% Another aspect of the classical description of an ARG data structure that
% derives from its historical origins as a stochastic
% process is how we encode information about genetic
% inheritance in the graph. The approach taken by Griffiths to
% doing this is elegant: all we need to do is store
% the location of the breakpoint at each recombination event.
% % Storing a breakpoint for each recombinatin event is a very
% % natural and straightforward thing to do when simulating
% % an  ARG process.
% Given the graph topology and these breakpoints, we can then
% recover the
% local trees at every point in the genome
% by tracing rootwards from the leaves and choosing the
% appropriate outbound path at recombination nodes.
% However, there are several significant disadvantages
% to this approach: [TODO be concrete here later
% and give pointers to the section where we discuss the
% alternative.]

% Fortunately, there is a straightforward change to the
% classical eARG encoding that simplifies
% implementations and greatly increases the flexibility
% of both the types and granularity of inheritance information
% that we can represent.
% The main change that we need to make is to associate
% the details of inheritance with \emph{edges} rather than
% as breakpoints associated with recombination events and
% edge ordering rules.
% This is illustrated in Fig~\ref{fig-arg-data-structure}C
% where this alternative edge annotation formulation is
% shown. By associating the interval $(0, x)$ with edge $(d,e)$
% and the interval $(x, L)$ with $(d, f)$ we make the encoding
% unambiguous: we can list the edges in any order, and it
% will describe the same set of local genealogical trees.
% This approach is also more flexible, because any pattern
% of inheritance can be described by these intervals. For
% example, if a gene conversion event occurred where a segment
% between positions $x$ and $y$ was inherited by
% node $d$ from $e$, we would state this by associating
% the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
% $\{(x, y]\}$ with $(d,f)$.
% This change is straightforward, involves no loss of information
% and (as we see in later sections)
% greatly increases both the expressivity and computational
% power of ancestral recombination graph data structures.

\section*{Genome ARGs}
As we have seen in the previous section, the focus
on a limited palette of events makes describing currently-available
datasets and known patterns of genetic inheritance as an eARG
cumbersome at best.
There is no fundamental reason that the patterns of genetic inheritance
that we wish to describe via an ARG must be in terms
of discrete \emph{events}, however. Recent discussions have illustrated the
relationship between ARGs and
the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
and the deep connections between the structures are
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}.
Following \cite{mathieson2020ancestry},
rather defining the inheritance
of genetic material in terms of \emph{events} as we do in an eARG,
we can instead define an ARG data structure in terms of
\emph{genomes} and intervals of genetic inheritance
along the edges.
To distinguish from the classical Griffiths eARG,
we refer to this encoding of an ARG data structure
as a ``Genome ARG'', or gARG.
The distinction between nodes representing events
and genomes may seem pedantic at first, but as we
see in later sections, we gain a great deal of flexibility
and computational efficiency by decoupling our representation
of genetic ancestry from the events that generated it.

@@

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.
Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity what can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome.
In large deeply sequenced pedigrees such complexities are commonplace.
Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can intepret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes.
Whether one regards nodes as genomes or events is orthogonal,
and something of a philosophical question which makes little
difference from a computational perspective.

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
parent to child (and ultimately from cell to cell) through some
number of generations. The edges are ``annotated''
with the genome coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to regard the annotations as describing
the complete set of genomic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.

@@

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to simulate or infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlikey to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
The penultimate local tree is highlighted to aid discussion.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the ``meaningful'' topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
edge annotations.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, \emph{all}
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, it removes potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

\section*{Precision of ancestry information}
% Not all nodes in an ARG are equal. Assuming that we have sequenced the
% genomes of the samples, there are intrinsic limits on what can
% be known about more ancient nodes, determined by the graph topology
% and the flow of ancestral material through it.

Ancestral recombination graphs are often characterised
as consisting of a complete record of the genetic history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information
will always be available, and in particular whether our
representation of genealogical history should \emph{require}
such precision. Simulations will usually generate complete
information about the history of a sample, of course, but
we may not have sufficient information to infer such detail.

@@

To illustrate this critical point, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
unrelated local trees
% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% for discussion. I'm sure there were other examples of people saying
% that
(as claimed by e.g.~\citet{hejase2020summary}),
and information about recombination and the correlation structure
between trees is not ``all or nothing''.
Both methods capture correlation structure between trees:
\relate\ first infers local tree topologies and then identifies edges
that persist from one tree to the next as a separate step
(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
\texttt{tsinfer} explicitly constructs nodes that are shared across local
trees.

@@
